Сплинты корневых систем для специальных подалгебр Ли

### 2015-03-23 AffineLieAlgebras/Splints/special-paper-ru.tex
Russian translation completed

For the adjoint representation of algebra $A_{r}$ we see that the projection $\Delta'$ of the root system
$\Delta_{\gf}$ should consist of roots of the subalgebra $\af=A_{r}$ with the multiplicity 2. The total
number of the roots for $A_{r}$ is $r(r+1)$, so $\Delta'$ and $\Delta_{\gf}$ contain $2r(r+1)$ roots,
since there are no roots orthogonal to $\Delta_{\af}$. There are two possible solutions for
$\gf$: $D_{r+1}$ and $G_{2}$ for $r=2$. But the dimension of $D_{r+1}$ is $(r+1)(2r+1)$ while the
dimension of $A_{r}$ is $r(r+2)$ and $2\mathrm{dim}A_{r}>\mathrm{dim}D_{r+1}$, so the adjoint
representation of $D_{r+1}$ cannot be decomposed as twice the adjoint representation of $A_{r}$.
There is no corresponding embedding $A_{r}\to D_{r+1}$ and no splint. We see that the conditions
\ref{eq:5}\ref{eq:6} do not hold.

Let's consider the embedding $B_{r}\to \gf$ such that the adjoint representation of $\gf$ is
decomposed into two adjoint representations of $B_{r}$. It is straightforward to check that there is
no solutions for $\gf$ satisfying the \ref{eq:5}\ref{eq:6} for the series $A$, $B$, $C$,
$D$ and all exceptional simple Lie algebras.

Since the number of roots of the root system $C_{r}$ is the same as for $B_{r}$ and we need to check the
same cases, we see that there is no embedding for $C_{r}$ too.

The adjoint representation of $G_{2}$ gives us the algebra $D_{4}$ as the solution of the
constraints \ref{eq:5}\ref{eq:6}. The embedding $G_{2}\to D_{4}$ exists, since there are
embeddings $G_{2}\to B_{3}$ and $B_{3}\to D_{4}$, but the decomposition of the adjoint representation of
$D_{4}$ is different: $ \mathrm{ad}_{D_{4}}=\mathrm{ad}_{G_{2}}\oplus 2 L^{\omega_{1}}_{G_{2}}$. So
this case does not produce splint for the projection $\Delta'$.

For the adjoint representation of $F_{4}$ there is no solution that satisfies conditions
\ref{eq:5}\ref{eq:6}.

@@

The complete classification of splints for special embeddings $\Delta'=\pia(\Delta_{\gf})
\equiv \Delta_{\af}\cup \Delta_{\sfr}$ where $\gf, \af$ are simple and $\sfr$ is semisimple is:
\begin{itemize}
\item $B_{n}\to D_{n+1}$
\item $G_{2}\to B_{3}$
%\item $G_{2}\to D_{4}$
\end{itemize}

Note that in the splints for special embeddings there is no case where $\Delta_{\sfr}$ is embedded
non-metrically which is a direct result of our exhaustive classification of suitable characteristic
representations of the subalgebra $\af$.

Having obtained the complete classification of splints of the projections of
the root systems for special embeddings, we need to check whether such splints lead
to the coincidence of branching coefficients with weight multiplicities in modules of
the algebra $\sfr$.

Let's consider the first case - the embedding $B_{n}\to D_{n+1}$. The root system of $D_{n+1}$ has
exactly two simple roots that are different from the simple roots of $B_{n}$. In the notation of
\cite{bourbaki2002lie} they are $\alpha_{n}=e_n-e_{n+1}$ and $\alpha_{n+1}=e_n+e_{n+1}$ while
$\alpha_n^{B_n}=e_n$. This difference leads to the crucial difference in the Weyl groups of the
algebras: while $W_{B_{n}}$ is a semidirect product of the group of permutation $e_{i} \to e_{j}$
and the group of the change of sign $e_{i}\to (\pm 1)_{i}e_{i}$, $W_{D_{n+1}}$ is the same but for
the additional condition on the group of the change of sign: $\prod_{i} (\pm)_{i}=1.$ Since the
projection of $D_{n+1}$ on $B_{n}$ acts as
$(e_1,e_2,\dots,e_{n},e_{n+1})\to(e_1,e_2,\dots,e_{n},0)$, the singular element
$\Phi^{(\mu)}_{D_{n+1}}$ which is the orbit of $W_{D_{n+1}}$ will become a composition of the orbits
of $W_{B_{n}}$ after the projection.

Indeed, if $\mu+\rho_{D_{n+1}}=(a_1,a_2,\dots,a_{n+1})$ in the standart basis
$(e_1,e_2,\dots,e_{n+1})$, then the Weyl group $W_{D_{n+1}}$ will act on it by permutating $a_i$ and
changing their signs. After the projection the last coefficient in the every element will be cut. Among
these elements will be groups in which all elements will have the same set of $a_i$ standing in
arbitrary order and having an arbitrary sign. These groups will be similar to the orbits of the Weyl
group $W_{B_{n}}$ but for the singular miltiplicity. A singular miltiplicity is a determinant
$\varepsilon (w)$ of the element $w$ of the Weyl group. Because of the additional condition on the
Weyl group $W_{D_{n+1}}$ the change of the signs of $a_i$ doesn't change the singular multiplicity.
That's not true for the Weyl group $W_{B_{n}}$, thus exactly half of the elements of each newly
formed orbit $\tilde\Phi^{\tilde\nu}_{B_n}$ of $W_{B_{n}}$ will have the wrong singular
multiplicity. Moreover, these orbits can be rewritten in the following form:

\begin{equation}
\tilde\Phi^{\tilde\nu}_{B_n}=\sum_{w\in W_{D_n}} \varepsilon(w) (1+s_{e_n})w e^{\tilde\nu+\rho_{B_n}}.
\end{equation}

This decomposition is possible because the orbit of the Weyl group of $D_n$ coincides with the half
of $\tilde\Phi^{\tilde\nu}_{B_n}$ that has correct singular multiplicities and other half can be
obtained by the action of the element $s_{e_n}$ of $W_{B_{n}}$.

Thus, the projection $(\Phi^{(\mu)}_{D_{n+1}})'$ consists of $(n+1)$ quasi-orbits (with ``wrong'' signs) of the Weyl group
$W_{B_{n}}$. It means that there are $(n+1)$ weights in the fundamental Weyl chamber $\bar C_{B_n}$
of $B_n$. It can be easily shown that these weights are $(a_1,a_2,\dots,a_{n})$ and weights obtained
from $(a_1,a_2,\dots,a_{n})$ by consequent subtraction of $\mu_i e_i$ starting with $\mu_n e_n$,
where $\mu_i$ is the Dynkin labels of $\mu$ plus $1$. As it turns out, if one were to add to this set
the required weights with the corresponding singular multiplicities to construct the singular element of
$\sfr=A_1+A_1+\dots+A_1$ as was done in the Introduction of this paper, all additional weights would
lie on the boundaries of the fundamental Weyl chamber $\bar C_{B_n}$. This fact is easily proved by
observing that the action of one of $s_{\alpha_i^{B_n}}$ on the weights doesn't change those
weigths. Thus, the projection $(\Phi^{\mu}_{D_{n+1}})'$ can be viewed as a quasi-orbit of the
singular element of $\sfr$:

\begin{equation}
\Phi^{[\mu_1,\mu_2,\dots,\mu_{n+1}]}_{D_{n+1}}=\sum_{w\in W_{D_n}} \varepsilon(w) (1+s_{e_n})w (e^{\mu'+\rho_{D_{n+1}}'-\tilde\mu}\Psi^{[\mu_1,\mu_2,\dots,\mu_{n}]}_{\sfr}),
\end{equation}
where $\tilde\mu=[\mu_1,\mu_2,\dots,\mu_{n}]$. This decomposition yields the branching rules similar to (\ref{bran1}):

\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{[\mu_1,\mu_2,\dots,\mu_{n+1}]}=M_{\left( \sfr\right) \widetilde{\nu }}^{[\mu_1,\mu_2,\dots,\mu_{n}]}, 
\label{bran11}
\end{equation}
which is in total agreement with the Gelfand-Tsetlin branching rules as the weights of the representations of $\sfr=A_1+A_1+\dots+A_1$ are always equal to one (see Fig.\ref{ris1} and Fig.\ref{ris2}).

\begin{figure}[h]
\begin{center}
\begin{minipage}[h]{0.5\linewidth}
\includegraphics[width=0.95\linewidth]{drawing-1}
\caption{Splint $D_3= B_2\cup A_1+A_1$: the projection of the singular element $\Phi_{D_3}^{[1,1,2]}$ can be made from singular elements $\Phi_{A_1+A_1}^{[1,1]}$. Light-grey area is the fundamental Weyl chamber of subalgebra $B_2$.} 
\label{ris1} 
\end{minipage}
\hfill
\begin{minipage}[h]{0.47\linewidth}
\includegraphics[width=0.9\linewidth,viewport=0 0 157 157,clip]{drawing-3}
\caption{The injection fan applied to the singular element $\Phi_{A_1+A_1}^{[1,1]}$ yields branching coefficients that are equal to 1.}
\label{ris2}
\end{minipage}
\end{center}
\end{figure}

A careful examination of the embedding $G_{2}\to B_{3}$ leads to exactly the same conclusion.

So far we only considered special splints where both embedded systems are root systems. However, it
is possible to study the splints where it is not the case. Such splints obviously can not serve to
simplify calculation of the branching rules but one may use the properties of special embeddings and
the projections to creat a repesentation theory for the systems that are not root systems.

As the result we see that branching coefficients coincide with weight multiplicities for the splints
in the following table:
\begin{equation}
\label{tab:2}
\begin{array}{cc||c|c}
\hbox{type} & \hspace{0.25in}\Delta \hspace{0.25in} & \hspace{0.25in}\Delta
_{\frak{a}}\hspace{0.25in} & \hspace{0.25in}\Delta _{\sfr}\hspace{0.25in}
\\ \hline\hline
\hbox{(i)} & G_{2} & A_{2} & A_{2} \\
& F_{4} & D_{4} & D_{4} \\ 
\hbox{(*)} & B_{3} & G_{2} & A_{2}  \\
\hbox{(*)} & D_{r+1} & B_{r} & \oplus ^{r}A_{1}  \\
\hline
\hbox{(ii)} & B_{r}(r\geq 2) & D_{r} & \oplus ^{r}A_{1} \\
\hbox{(iii)} & A_{r}(r\geq 2) & A_{r-1}\oplus u\left( 1\right)  & \oplus
^{r}A_{1} \\
& B_{2} & A_{1}\oplus u\left( 1\right)  & A_{2}
\end{array}
\end{equation} 
here the splints marked with (*) are special splints.

\section{Conclusion}
\label{sec:conclusion}

The computation of branching coefficients is important for different physical models with a symmetry
breaking. This computation is drastically simplified if branching coefficients coincide with weight
multiplicities, since efficient Freudental formula can be used \cite{moody1982fast}. The classification
of splints for regular and special embeddings gives us all the cases when this coincidence takes
place. Aside from computational importance, this coincidence is very interesting from
representation-theoretic point of view, because it follows from the new unexpected connection between
different simple Lie algebras.

There exist special embeddings of semisimple Lie algebras into simple Lie algebras. For such
embeddings the procedure of finding splints that allow to calculate branching coefficient is similar
to the one conducted above. While the results of the procedure are not presented in this paper they
can be obtained with little difficulty.

\section*{Acknowledgements}
\label{sec:acknowledgements}

We thank the organizers of the conference ``In Search of Fundamental Symmetries'' where some results
of this work were presented.

P.I. Kakin would like to acknowledge Saint Petersburg State University for the research grant
11.38.185.2014.

>>>>

Рассмотрим присоединенное представление алгебры $A_{r}$. Мы видим, что проекция $\Delta'$ корневой
системы $\Delta_{\gf}$ должна состоять из корней подалгебры $\af=A_{r}$ с кратностью 2. Общее число
корней в $A_{r}$ равно $r(r+1)$, так что $\Delta'$ и $\Delta_{\gf}$ содержат по $2r(r+1)$ корней,
так как нет корней, ортогональных $\Delta_{\af}$. Отсюда существует два варианта решения для алгебры
$\gf$: $D_{r+1}$ и $G_{2}$ при $r=2$. Но размерность $D_{r+1}$ равна $(r+1)(2r+1)$, тогда как
размерность $A_{r}$ равна $r(r+2)$ и $2\mathrm{dim}A_{r}>\mathrm{dim}D_{r+1}$, так что
присоединенное представление $D_{r+1}$ не может быть разложено в сумму двух копий присоединенного
представления $A_{r}$. А значит, нет и соответствующего вложения $A_{r}\to D_{r+1}$ и сплинта.
Условия \ref{eq:5},\ref{eq:6} не выполняются.

Теперь перейдем к вложению $B_{r}\to \gf$, такому что присоединенное представление $\gf$
раскладывается в сумму двух копий присоединенного представления $B_{r}$. Легко проверить перебором,
что подходящего решения для $\gf$, удовлетворяющего условиям \ref{eq:5}\ref{eq:6}, не существует в
сериях $A$, $B$, $C$, $D$, не подходят и все исключительные простые алгебры Ли.

Так как число корней корневой системы  $C_{r}$ такое же, как и в $B_{r}$ и нам нужно проверять те же
случаи, мы видим, что для  $C_{r}$ тоже нет подходящего вложения.

Присоединенное представление алгебры $G_{2}$ дает нам алгебру  $D_{4}$ в качестве решения условий
\ref{eq:5},\ref{eq:6}. Вложение  $G_{2}\to D_{4}$ существует, так как есть вложения $G_{2}\to B_{3}$
и $B_{3}\to D_{4}$, но разложение присоединенного представления $D_{4}$ другое: $
\mathrm{ad}_{D_{4}}=\mathrm{ad}_{G_{2}}\oplus 2 L^{\omega_{1}}_{G_{2}}$. То есть и этот случай не
приводит к возникновению сплинта для проекции $\Delta'$.

Для присоединенного представления  $F_{4}$ решений, удовлетворяющих соотношениям
\ref{eq:5},\ref{eq:6} не существует.

@@

Итак, полная классификация сплинтов для специальных вложений $\Delta'=\pia(\Delta_{\gf})
\equiv \Delta_{\af}\cup \Delta_{\sfr}$, где  $\gf, \af$ -- простые, а  $\sfr$ -- полупростая алгебра
Ли имеет следующий вид:
\begin{itemize}
\item $B_{n}\to D_{n+1}$
\item $G_{2}\to B_{3}$
%\item $G_{2}\to D_{4}$
\end{itemize}

Заметим, что среди сплинтов для специальных вложений нет случаев, когда $\Delta_{\sfr}$ вложена не
метрически. Это результат нашей исчерпывающей классификации подходящих характеристических
представлений подалгебры $\af$.

После получения полной классификации сплинтов проекций корневых систем при специальных вложениях,
необходимо проверить совпадение коэффициентов ветвления с кратностями весов в модулях алгебры $\sfr$.

Рассмотрим первый случай -- вложение $B_{n}\to D_{n+1}$. Корневая система $D_{n+1}$ содержит в
точности два простых корня, не входящих в число простых корней подалгебры $B_{n}$. В обозначениях Бурбаки
\cite{bourbaki2002lie} это простые корни $\alpha_{n}=e_n-e_{n+1}$ и $\alpha_{n+1}=e_n+e_{n+1}$, а
$\alpha_n^{B_n}=e_n$. Такое отличие приводит к критическому различию в группах Вейля этих алгебр:
группа $W_{B_{n}}$ представляет собой полупрямое произведение группы перестановок $e_{i} \to e_{j}$
и группы смен знаков $e_{i}\to (\pm 1)_{i}e_{i}$, а группа $W_{D_{n+1}}$ отличается тем, что на
операции смены знаков наложено дополнительное условие: $\prod_{i} (\pm)_{i}=1.$ Так как проекция
$D_{n+1}$ на $B_{n}$ действует как
$(e_1,e_2,\dots,e_{n},e_{n+1})\to(e_1,e_2,\dots,e_{n},0)$, сингулярный элемент
$\Phi^{(\mu)}_{D_{n+1}}$, представляющий собой орбиту $W_{D_{n+1}}$, раскладывается в комбинацию
орбит  $W_{B_{n}}$ после проекции.

А именно, если $\mu+\rho_{D_{n+1}}=(a_1,a_2,\dots,a_{n+1})$ в стандартном базисе
$(e_1,e_2,\dots,e_{n+1})$, то группа Вейля $W_{D_{n+1}}$ действует на него перестановкой координат
$a_i$ и сменой их знаков. После проектирования последняя координата каждого веса в орбите должна быть
отброшена. Среди элементов орбиты будут подмножества, в которых все элементы имеют один и тот же
набор координат $a_i$, но стоящих в произвольном порядке и имеющих произвольные знаки. Эти
подмножества похожи на орбиты группы Вейля $W_{B_{n}}$, но с сингулярными кратностями. Сингулярная
кратность -- это детерминант
$\varepsilon (w)$ элемента  $w$ группы Вейля. В силу дополнительного условия в группе Вейля
$W_{D_{n+1}}$ смена знаков  $a_i$ не меняет сингулярной кратности. Однако это не выполняется в
группе Вейля $W_{B_{n}}$, а значит ровно половина элементов в каждой новопостроенной орбите
$\tilde\Phi^{\tilde\nu}_{B_n}$ группы $W_{B_{n}}$ будет с неправильной сингулярной кратностью. Более
того, эти орбиты можно переписать в следующем виде:

\begin{equation}
\tilde\Phi^{\tilde\nu}_{B_n}=\sum_{w\in W_{D_n}} \varepsilon(w) (1+s_{e_n})w e^{\tilde\nu+\rho_{B_n}}.
\end{equation}
Это разложение возможно потому, что орбита группы Вейля $D_n$ совпадает с той половиной
$\tilde\Phi^{\tilde\nu}_{B_n}$, у которой сингулярные кратности правильные, а вторая половина
получается действием элемента $s_{e_n}$ группы $W_{B_{n}}$.

Итак, проекция $(\Phi^{(\mu)}_{D_{n+1}})'$ состоит из  $(n+1)$ квази-орбиты (с  ``неправильными''
знаками) группы Вейля$W_{B_{n}}$. Значит, в главной камере Вейля  $\bar C_{B_n}$ алгебры
$B_n$ содержится
$(n+1)$ вес.  Легко проверить, что это старший вес  $(a_1,a_2,\dots,a_{n})$ и веса, получающиеся из
него последовательными вычитаниями $\mu_i e_i$, начиная с $\mu_n e_n$, где  $\mu_i$ -- индексы
Дынкина  $\mu$ плюс $1$. Оказывается, что при дополнении этого множества  весами с
соответствующими сингулярными кратностями до восстановления сингулярного элемента
$\sfr=A_1+A_1+\dots+A_1$, как было сделано во введении, все дополнительные веса будут лежать на
границе главной камеры Вейля $\bar C_{B_n}$. Этот факт легко доказать, если заметить, что действие
одного из отражений of $s_{\alpha_i^{B_n}}$ на веса не меняет их. Итак, проекция
$(\Phi^{\mu}_{D_{n+1}})'$ может рассматриваться как квази-орбита сингулярного элемента $\sfr$:

\begin{equation}
\Phi^{[\mu_1,\mu_2,\dots,\mu_{n+1}]}_{D_{n+1}}=\sum_{w\in W_{D_n}} \varepsilon(w) (1+s_{e_n})w (e^{\mu'+\rho_{D_{n+1}}'-\tilde\mu}\Psi^{[\mu_1,\mu_2,\dots,\mu_{n}]}_{\sfr}),
\end{equation}
где $\tilde\mu=[\mu_1,\mu_2,\dots,\mu_{n}]$. Это порождает  правила ветвления, аналогичные соотношению (\ref{bran1}):

\begin{equation}
b_{\left( \mu -\phi \left( \widetilde{\mu }-\widetilde{\nu }\right) \right)
}^{[\mu_1,\mu_2,\dots,\mu_{n+1}]}=M_{\left( \sfr\right) \widetilde{\nu }}^{[\mu_1,\mu_2,\dots,\mu_{n}]}, 
\label{bran11}
\end{equation}
что полностью соответствует правилам ветвления Гельфанда-Цетлина, так как кратности весов в
представлениях  $\sfr=A_1+A_1+\dots+A_1$ всегда равны единице (см Рис.\ref{ris1} и Рис.\ref{ris2}).

\begin{figure}[h]
\begin{center}
\begin{minipage}[h]{0.5\linewidth}
\includegraphics[width=0.95\linewidth]{drawing-1}
\caption{Сплинт $D_3= B_2\cup A_1+A_1$: проекция сингулярного элемента  $\Phi_{D_3}^{[1,1,2]}$ может
  быть построена из сингулярных элементова $\Phi_{A_1+A_1}^{[1,1]}$. Светло-серая область -- это
  главная камера Вейля подалгебры $B_2$.} 
\label{ris1} 
\end{minipage}
\hfill
\begin{minipage}[h]{0.47\linewidth}
\includegraphics[width=0.9\linewidth,viewport=0 0 157 157,clip]{drawing-3}
\caption{Веер вложения, примененный к сингулярному элементу $\Phi_{A_1+A_1}^{[1,1]}$, дает
  коэффициенты ветвления, равные  1.}
\label{ris2}
\end{minipage}
\end{center}
\end{figure}

Аналогичное детальное рассмотрение вложения $G_{2}\to B_{3}$ дает такой же результат.

Мы рассмотрели случаи сплинтов для специальных вложений, в которых вложенные системы являются
корневыми. Однако можно изучать сплинты и тогда, когда это не выполняется. Такие сплинты нельзя
использовать для построения правил ветвления, но можно надеяться использовать свойства специальных
вложений и проекций, чтобы построить теорию представлений для систем векторов, не являющихся
корневыми системами.

В результате мы видим, что коэффициенты ветвления совпадают с кратностями весов для сплинтов из
следующей таблицы:
\begin{equation}
\label{tab:2}
\begin{array}{cc||c|c}
\hbox{type} & \hspace{0.25in}\Delta \hspace{0.25in} & \hspace{0.25in}\Delta
_{\frak{a}}\hspace{0.25in} & \hspace{0.25in}\Delta _{\sfr}\hspace{0.25in}
\\ \hline\hline
\hbox{(i)} & G_{2} & A_{2} & A_{2} \\
& F_{4} & D_{4} & D_{4} \\ 
\hbox{(*)} & B_{3} & G_{2} & A_{2}  \\
\hbox{(*)} & D_{r+1} & B_{r} & \oplus ^{r}A_{1}  \\
\hline
\hbox{(ii)} & B_{r}(r\geq 2) & D_{r} & \oplus ^{r}A_{1} \\
\hbox{(iii)} & A_{r}(r\geq 2) & A_{r-1}\oplus u\left( 1\right)  & \oplus
^{r}A_{1} \\
& B_{2} & A_{1}\oplus u\left( 1\right)  & A_{2}
\end{array}
\end{equation} 
Здесь звездочкой (*) отмечены сплинты, возникающие в специальных вложениях.

\section{Заключение}
\label{sec:conclusion}

Вычисление коэффициентов ветвление требуется в различных физических моделях с нарушением симметрии.
Такое вычисление радикально упрощается, если коэффициенты ветвления совпадают с кратностями весов,
так как для них можно использовать эффективную формулу Фрейденталя
\cite{moody1982fast}.Классификация сплинтов для регулярных и специальных вложений дает нам все
случаи такого совпадения. Помимо облегчения вычислений это совпадение интересно с точки зрения
теории представлений, так как оно указывает на новую и неожиданную связь между различными алгебрами Ли.

Существуют также специальные вложения полупростых алгебр Ли в простые алгебры Ли. Для таких вложений
процедура нахождения сплинтов, позволяющих вычислять коэффициенты ветвления, аналогична
рассмотренной выше. Хотя мы не приводим результаты этой процедуры, их нетрудно получить.

\section*{Благодарности}
\label{sec:acknowledgements}

Мы благодарим организаторов конференции ``In Search of Fundamental Symmetries'', где были
представлены некоторые результаты данной работы.

П.И. Какинь выражает благодарность СПБГУ за финансовую поддержку в рамках исследовательского гранта  11.38.185.2014.
